---
title: "durumGEAr: A Confound-Aware GEA Modelling Workflow"
author: "Sabri Youssef"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{durumGEAr: A Confound-Aware GEA Modelling Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5
)
```

```{r cache-load, include = FALSE}
## This vignette displays REAL COMPUTED NUMBERS, not eval=FALSE stubs
## (adversarial-review issue #1). The cross-validation and permutation-test
## chunks below (spatialBlockCV, naiveCV, locoCV, confoundGapTest,
## residualize, driverAnalysis, scoreThenCluster) are individually cheap to
## call, but the FULL set of them together (in particular confoundGapTest's
## and residualize's permutation loops) takes on the order of 30-40 minutes
## even at reduced settings. Re-running all of that on every `R CMD build`
## is impractical, so the numbers are pre-computed once by
## `data-raw/build_vignette_cache.R` (using deliberately lighter settings
## than the package defaults -- fewer seeds/trees/permutations -- purely to
## keep the vignette build tractable) and cached to
## `inst/extdata/vignette_cache.rds`. Every code chunk below is still the
## exact call that produced the cached object; only the *execution* is
## replaced by a cache lookup, and the object printed is identical to what
## calling the function live would return.
cache_path <- system.file("extdata", "vignette_cache.rds", package = "durumGEAr")
if (!nzchar(cache_path)) {
  # package not yet installed (e.g. building the vignette in place during
  # development) -- fall back to the source-tree copy
  cache_path <- "../inst/extdata/vignette_cache.rds"
}
vc <- readRDS(cache_path)
```

## Why this package exists

Genotype-environment association (GEA) data from genebank collections is
routinely **pseudoreplicated** (many accessions share one collection site, so
their bioclimatic predictors are literally identical) and **spatially
confounded** (country of origin can stand in for climate, inflating apparent
predictive skill). A naive interpolation R² on such data is not evidence of a
transferable environment-to-genetics relationship — it may just be country
memorization. `durumGEAr` packages the full defensible workflow used to tell
these apart, on a durum wheat (*Triticum turgidum* ssp. *durum*) genebank
collection, following the function-naming conventions of `icardaFIGSr`.

The pipeline has four stages: (1) collapse pseudoreplicated rows to
independent effective units, (2) evaluate with spatially-aware
cross-validation that never lets one site or one country appear in both
train and test, (3) residualize out country identity to see what climate
signal remains, and (4) run a permutation-based driver analysis on what's
left.

```{r setup}
library(durumGEAr)
```

## 1. Collapsing pseudoreplicated rows to effective units

```{r}
data(durumRaw)
dim(durumRaw)

cu <- collapseUnits(durumRaw)
print(cu)
```

`cu$audit` is the full cleaning ledger: raw rows, exact duplicates, cloned
target vectors, IG duplicates, minimum admixture-membership confidence, and
the resulting confident-unit count. `cu$confident` is the canonical modelling
frame — one row per (SiteCode x Cluster) effective unit, restricted to units
with a clean genetic mean and confident admixture assignment. It matches the
packaged `durumUnits` dataset:

```{r}
data(durumUnits)
dim(durumUnits)
stopifnot(nrow(cu$confident) == nrow(durumUnits))
```

## 2. Environmental PCA (ePC)

`computeEPC()` rotates the 19 bioclimatic variables (+ altitude) into a small
set of orthogonal axes:

```{r}
ep <- computeEPC(durumUnits)
round(ep$var_explained, 3)
```

## 3. Spatially-aware cross-validation

**Interpolation** (grouped by SiteCode, so no site straddles a fold):

```{r}
## sb <- spatialBlockCV(durumUnits, seeds = 42:44, num.trees = 300, min.node.size = 3)
## (cached; see "cache-load" chunk above -- lighter settings than the
##  num.trees = 600 package default, purely to keep the vignette build fast)
sb <- vc$sb
sb
```

**A leaky baseline for comparison** — `naiveCV()` (adversarial-review issue
#2) runs ordinary, *ungrouped* random k-fold CV, so accessions from the same
site can appear on both sides of a fold. It quantifies how much apparent
skill the site-level pseudoreplication alone is worth:

```{r}
## nv <- naiveCV(durumUnits, seeds = 42:44, num.trees = 300, min.node.size = 3)
nv <- vc$naive
nv
```

Here the naive, leaky mean R2 (`r round(vc$naive$metrics$mean_R2, 3)`) is
actually *lower* than the properly grouped `spatialBlockCV` mean
(`r round(vc$sb$metrics$mean_R2, 3)`) for this particular dataset and fold
draw -- a reminder that pseudoreplication does not always inflate a metric
in the naive direction; what matters is that `spatialBlockCV` is the
methodologically correct comparison, not that it is always the higher
number.

**Extrapolation stress tests** — leave-one-country-out and leave-one-cluster-out:

```{r}
## lo <- locoCV(durumUnits, group = "Country", num.trees = 150, min.node.size = 3)
lo <- vc$loco
lo$metrics$mean_R2
lo$median_group_R2
nrow(lo$per_group)              # number of countries with >= 2 members
sum(lo$per_group$beats_baseline)   # how many countries beat a no-skill baseline

## lc <- locoCV(durumUnits, group = "Cluster", num.trees = 150, min.node.size = 3)
lc <- vc$loclu
lc$metrics$mean_R2
lc$per_group
```

The gap between the interpolation number
(`r round(vc$sb$metrics$mean_R2, 3)`) and the leave-one-country-out number
(`r round(vc$loco$metrics$mean_R2, 3)`) is the diagnostic for spatial confounding: a
model that only memorizes country identity collapses under
`locoCV(group = "Country")` even though it looks strong under
`spatialBlockCV()`. Leave-one-cluster-out is even harsher
(mean R2 = `r round(vc$loclu$metrics$mean_R2, 3)`) because whole genetic clusters are
also geographically clustered.

**Is that gap real, or could it arise by chance from a small-country /
sparse-data effect?** `confoundGapTest()` (adversarial-review issue #3)
answers this directly: it holds the interpolation estimate fixed, builds a
null distribution for the LOCO gap by repeatedly shuffling *Country* labels
and re-running `locoCV()` with identical hyperparameters, and reports a
one-sided p-value.

```{r}
## gt <- confoundGapTest(durumUnits, seed = 42, n_perm = 60,
##                        num.trees = 15, min.node.size = 3)
## (n_perm and num.trees are reduced from the package defaults of 100/600
##  purely for vignette build time; see ?confoundGapTest "Expected runtime")
gt <- vc$gaptest
gt
```

The observed gap (`r round(vc$gaptest$observed_gap, 3)`) sits well above the
permutation null (mean `r round(vc$gaptest$null_mean, 3)`,
SD `r round(vc$gaptest$null_sd, 3)`), giving p = `r round(vc$gaptest$p_value, 4)`
at `r vc$gaptest$n_perm` permutations. That p-value sits exactly at this run's
resolution floor (`1/(n_perm+1)` = `r round(1/(vc$gaptest$n_perm+1), 4)`, printed
above as a WARNING) -- with only 60 permutations, the smallest p-value this test
can ever report is 0.0164, so this result tells us the gap is not an artefact of
small-country noise *at this coarse resolution*, but does not pin down how much
smaller the true p-value might be. The package default of `n_perm = 100` (and a
higher `num.trees`, both reduced here purely for vignette build time; see
`?confoundGapTest` "Expected runtime") gives a finer-grained, floor-free estimate.

## 4. Residualizing out country identity

`residualize()` asks a stricter question per target: after removing each
fold's country-mean fold-safe, is there still climate signal in the
residual? The verdict is no longer a hardcoded R2 cutoff
(adversarial-review issue #6) but a within-country permutation significance
test: the target is shuffled *inside* each country (preserving country
structure and group sizes), the fold-safe residual fit is rerun, and the
verdict is `"real within-country signal"` only if the observed residual R2
beats that null at `alpha = 0.05`.

```{r}
## rz <- residualize(durumUnits, seeds = 42:43, n_perm = 100,
##                    num.trees = 50, min.node.size = 3)
## (num.trees reduced from the default of 600 purely for vignette build
##  time; see ?residualize "Expected runtime")
rz <- vc$rz
rz
```

```{r}
data.frame(target = names(rz$obs_R2_test),
           obs_R2_test = round(rz$obs_R2_test, 3),
           per_target_R2 = round(rz$per_target_R2, 3),
           p_value = round(rz$p_value, 3),
           verdict = rz$verdict,
           row.names = NULL)
```

The verdict is driven by `obs_R2_test` (the single-seed statistic actually
compared against the permutation null), shown alongside the multi-seed-averaged
`per_target_R2` headline number for context -- the two can disagree in sign for
a target sitting right around zero, which is exactly what happens here for
gPC1 (`obs_R2_test` = `r round(rz$obs_R2_test["gPC1"], 3)` vs. `per_target_R2` =
`r round(rz$per_target_R2["gPC1"], 3)`): its verdict of
`r rz$verdict["gPC1"]` should be read as fragile, not a confident finding, and
would need more seeds and a higher `n_perm` to resolve either way. gPC2 and
gPC5 give a robust `"real within-country signal"` verdict (both statistics
agree in sign and magnitude), while gPC3 and gPC4 are consistently
`"country-identity artifact"`. All p-values in this cached run sit at the
resolution floor (`r round(1/(rz$n_perm+1), 4)`, printed above as a WARNING) --
they establish that no target's residual signal is implausibly large relative
to a within-country-shuffled null at this resolution, not a precise significance
level. gPC2 stands out with a much larger residual R2
(`r round(rz$per_target_R2["gPC2"], 3)`) than the rest, which is why it is used
as the worked example for driver analysis below.

## 5. Driver analysis

For a target found to carry real residual signal, `driverAnalysis()` runs a
permutation test to rank which bioclimatic variables drive it. The null is
now a **max-statistic** null over *every* candidate predictor
(adversarial-review issue #4: comparing the single best-of-many predictor
against a null built only for that one predictor is a multiple-comparisons
problem -- see Westfall & Young 1993), and the observed statistic and every
null permutation use identical hyperparameters (issue #5: the observed fit
previously used `num.trees = 300` while the null used `num.trees = 150`).

```{r}
## dr <- driverAnalysis(durumUnits, "gPC2", seeds = 42, n_perm = 40,
##                       num.trees = 150, min.node.size = 5)
## (n_perm and num.trees reduced from defaults of 200/300; see
##  ?driverAnalysis "Expected runtime")
dr <- vc$dr_gpc2
head(dr$univariate)
dr$p_value
dr$n_predictors_tested
```

The top-ranked predictor's univariate R2 (`r round(dr$observed_R2, 3)`) is
judged against the max over all `r dr$n_predictors_tested` candidate
predictors under permutation (null mean
`r round(mean(dr$null_max), 3)`, SD `r round(sd(dr$null_max), 3)`), giving
p = `r dr$p_value` at `r dr$n_perm` permutations. At this reduced
permutation count for gPC2 the top univariate predictor does not clear the
corrected max-statistic null -- an honest negative result reported as-is,
illustrating exactly why the max-statistic correction (issue #4) matters:
the naive single-predictor null would have overstated significance here.

## 6. Stage-2: assigning accessions to genetic clusters

The regression above predicts the *continuous* genetic scores. A common
downstream question is discrete: which genetic **cluster** does an accession
belong to? `scoreThenCluster()` answers this with a leakage-free two-stage
procedure — Stage 1 predicts gPC1–gPC5 from environment with the same
spatial-block folds, and Stage 2 fits a quadratic discriminant classifier
that is trained **only on the true genetic scores of the training fold** and
then applied to the *predicted* scores of the held-out fold.

```{r}
## s2 <- scoreThenCluster(durumUnits, per_tree = FALSE, cv_ceiling = TRUE,
##                         num.trees = 300, min.node.size = 3)
s2 <- vc$s2
s2$accuracy      # honest out-of-fold cluster accuracy
s2$ceiling       # accuracy a regressor could reach, per ceiling_type below
s2$ceiling_type  # which ceiling estimator was used
s2$gap           # ceiling - accuracy = the cost of Stage-1 regression error
s2$confusion
```

By default `ceiling` is computed **in-sample** (QDA fit and evaluated on all
units at once), which is an optimistic upper bound rather than a validated
estimate -- some of the `gap` could be ceiling-optimism itself rather than
genuine Stage-1 error (adversarial-review issue #7). Passing
`cv_ceiling = TRUE`, as used to build the cached object above, instead fits
QDA **out-of-fold** on the same spatial-block folds used for `accuracy`, so
the printed ceiling (`r round(s2$ceiling, 3)`,
ceiling_type = "`r s2$ceiling_type`") is itself a validated, non-leaking
estimate -- at the cost of `k` extra QDA fits. The honest out-of-fold
accuracy (`r round(s2$accuracy, 3)`) sits `r round(s2$gap, 3)` below this
validated ceiling.

Two calibration diagnostics accompany the accuracy: posterior-entropy
quartiles (`s2$entropy_calibration`, accuracy should decline monotonically
as entropy rises) and an off-manifold typicality flag (`s2$off_manifold`)
for accessions whose predicted scores fall outside every cluster's 95%
Mahalanobis ellipsoid.

```{r}
s2$entropy_calibration
s2$state_accuracy
sum(s2$off_manifold)
```

Accuracy falls from `r round(s2$entropy_calibration$accuracy[1], 3)` in the
most-confident posterior-entropy quartile to
`r round(s2$entropy_calibration$accuracy[4], 3)` in the least-confident
quartile -- the classifier's own uncertainty is informative, not just
decorative. `r sum(s2$off_manifold)` of `r nrow(durumUnits)` units are
flagged off-manifold (predicted scores outside every cluster's 95%
Mahalanobis ellipsoid); their accuracy
(`r round(s2$state_accuracy$accuracy[s2$state_accuracy$state == "off_manifold"], 3)`)
is correspondingly the lowest of the three confidence states.

## 8. Per-target conclusions

Putting the diagnostics above together, target by target (all numbers from
the reduced-setting cache used to build this vignette; see individual
sections for the corresponding package defaults):

```{r, echo = FALSE, results = "asis"}
targets <- names(rz$per_target_R2)
for (tg in targets) {
  cat(sprintf(
    "- **%s**: interpolation R2 = %.3f (spatial-block); within-country residual R2 = %.3f (p = %.3f) -> *%s*.\n",
    tg, vc$sb$metrics$per_target$R2[vc$sb$metrics$per_target$target == tg],
    rz$per_target_R2[tg], rz$p_value[tg], rz$verdict[tg]
  ))
}
```

Every target in this run clears the within-country permutation significance
threshold, but the *size* of that residual signal varies enormously: gPC2
carries by far the largest fold-safe within-country R2
(`r round(rz$per_target_R2["gPC2"], 3)`), while the rest are small in
absolute terms even where "significant". Read the per-target residual R2,
not just the verdict label, before treating any target as strongly
climate-predictable within a country.

## 9. Mapping accessions

```{r, fig.alt = "Map of accession locations coloured by genetic cluster"}
mapAccessions(durumUnits, color_by = "Cluster",
              main = "Durum wheat effective units")
```

## 10. Stability under perturbation

A verdict is only useful if it survives reasonable changes to the
hyperparameters that produced it. The sensitivity study
(`inst/sensitivity_analysis.R`) reruns `confoundGapTest()` and `residualize()`
across a grid of `num.trees`, `min.node.size`, and `k`, and records, for each
target, how many *distinct* residualization verdicts appear across the grid
(1 = perfectly stable; >1 = the verdict flips somewhere in the grid).

```{r stability, echo = FALSE, results = "asis"}
if (!is.null(vc$sensitivity)) {
  se <- vc$sensitivity
  cat(sprintf(
    "Across %d grid cells, the observed confounding gap ranged from %.3f to %.3f (median %.3f). ",
    se$n_cells, se$gap_range[1], se$gap_range[2], stats::median(se$grid$observed_gap)))
  flips <- se$verdict_distinct_per_target
  stable <- names(flips)[flips == 1]
  unstable <- names(flips)[flips > 1]
  cat(sprintf("Verdict stability: %s held a single verdict across every cell",
              if (length(stable)) paste(stable, collapse = ", ") else "no target"))
  if (length(unstable))
    cat(sprintf("; %s flipped across the grid and should be treated as borderline.",
                paste(unstable, collapse = ", "))) else
    cat(" — no target flipped.")
  cat("\n")
} else {
  cat("*The sensitivity-grid cache is not bundled in this build. ",
      "Run `Rscript inst/sensitivity_analysis.R` (env vars `DURUM_SENS_*` set the ",
      "grid) to generate `inst/cache_sensitivity_grid.rds`, then rebuild the ",
      "vignette cache.*\n", sep = "")
}
```

These numbers come from a deliberately reduced grid so the study runs on a
laptop; the script accepts environment variables (`DURUM_SENS_NTREES`,
`DURUM_SENS_NODESIZE`, `DURUM_SENS_K`, `DURUM_SENS_NPERM`) so you can widen the
grid and raise `n_perm` for a publication-grade stability check on your own
hardware. The robust, quantitative takeaway is the **gap**: the
interpolation-vs-LOCO confounding gap stays large and positive in every cell
(here roughly 0.34–0.36), so the headline "most predictive skill is country
identity" conclusion does not depend on a lucky hyperparameter choice. Note that
the *per-target residualization verdicts* in this grid are computed at a very low
`n_perm` (the p-value floor `1/(n_perm+1)` is well above alpha), so no target can
reach a "real signal" verdict here regardless of its residual-scale R²; the
residual R² values themselves (e.g. gPC2 ≈ 0.18) still separate the signal-carrying
targets, and the significance verdict is properly powered only when `residualize()`
is rerun at production `n_perm` (see the study script and Section 11).

## 11. Permutation-test calibration

The verdicts rest on permutation p-values, so it is worth checking that the
permutation machinery is itself well-calibrated — i.e. that under a true null it
rejects at about the nominal rate rather than systematically over- or
under-rejecting. `inst/validate_permutation_tests.R` measures the empirical
Type I error two ways: for `residualize()` it shuffles each target *within
country* (destroying any within-country signal while preserving country
structure) and counts how often the test still calls "real signal"; for
`confoundGapTest()` it shuffles the country labels and counts spurious
significant gaps.

```{r calibration, echo = FALSE, results = "asis"}
if (!is.null(vc$calibration)) {
  ca <- vc$calibration
  cat(sprintf(
    "At the reduced study settings (residualize: N_REP=%d, N_PERM=%d, num.trees=%d; gap: N_REP=%d, N_PERM=%d, num.trees=%d), the empirical Type I error was %.3f for `residualize()` and %.3f for `confoundGapTest()`, against a nominal alpha of %.2f. ",
    ca$settings$N_REP, ca$settings$N_PERM, ca$settings$N_TREES,
    ca$settings$N_REP_GAP, ca$settings$N_PERM_GAP, ca$settings$N_TREES_GAP,
    ca$type1_resid_overall, ca$type1_gap, ca$settings$ALPHA))
  cat("A rate near alpha indicates the permutation test is calibrated; a rate ",
      "well above alpha would warn that the null is too easy to beat at these ",
      "settings.\n", sep = "")
} else {
  cat("*The permutation-calibration cache is not bundled in this build. ",
      "Run `Rscript inst/validate_permutation_tests.R` (env vars `DURUM_CAL_*` ",
      "set replicate/permutation counts) to generate ",
      "`inst/cache_permutation_calibration.rds`, then rebuild the vignette cache.*\n",
      sep = "")
}
```

As with the sensitivity grid, the shipped numbers are from a reduced run
(fewer replicates and permutations than you would use for a formal calibration
claim); the script is parameterized by environment variables so the same
measurement can be rerun at production settings. The p-value CIs reported by
`residualize()` and `confoundGapTest()` (v1.2.0) are bootstrapped from the
`n_perm`-row permutation null, so their resolution is capped by `n_perm`: at the
default `n_perm = 100` a p-value CI cannot be finer than roughly ±0.01. Read a
floor p-value (`1/(n_perm+1)`) as "beat every permutation at this resolution",
not as a finely estimated significance level.

## 12. Using the model in practice

The diagnostics above tell you *which* genetic axes carry transferable signal.
The deployment layer turns that judgement into predictions you can act on —
while keeping the reliability verdict attached to every number, so an
untrustworthy axis cannot be quietly presented as a confident prediction.

The reliability table is inherited directly from `residualize()`. Rather than
re-run that permutation loop here, we load the shipped
`durum_residualize_results` object (a worked `residualize()` result on
`durumUnits`) and hand it to `fitGeneticScoreModel()`, which fits one
ExtraTrees regressor per target and gates the axes flagged fragile.

```{r deploy-fit}
data(durum_residualize_results)
mod <- fitGeneticScoreModel(
  durumUnits,
  residualize_result = durum_residualize_results,
  num.trees = 100                    # light for the vignette; default is 300
)
mod
```

The printout lists, per axis, the `residualize()` verdict, its fragility grade,
the residual-scale R², and the resulting reliability label. Gating is
deliberately conservative: an axis is trusted only when its fragility grade
reaches `robust`. At the shipped object's reduced `n_perm = 30`, no axis can
reach `robust` — the permutation p-value floor (`1/(n_perm+1)`) sits above the
significance threshold — so **every axis is gated**. This is the honest default:
the package refuses to certify an axis the shipped validation cannot support.
`predict()` still returns all columns, but marks each so a caller never mistakes
an uncertified axis for a confirmed climate signal.

```{r deploy-predict}
newx <- durumUnits[1:5, ]
pr   <- predict(mod, newx)
round(pr, 3)
attr(pr, "gated_targets")           # axes the shipped validation does not certify
```

Pass `gated_to_na = TRUE` to blank the untrustworthy columns entirely. To
collapse axes into a single robust score, `robustGeneticScore()` drops the gated
axes by default — but here that would leave nothing, so it errors unless you
supply `keep=`. A production `residualize()` run (`n_perm = 100`,
`num.trees = 600`) is what promotes gPC2 (residual R² ≈ 0.18) and gPC5 (≈ 0.06)
to trusted; if you have run it and confirmed those axes, force-keep them:

```{r deploy-robust}
rs <- robustGeneticScore(mod, newx, keep = c("gPC2", "gPC5"))
data.frame(robust_score = round(rs, 3),
           axes = paste(attr(rs, "axes"), collapse = ", "))
```

Nothing is hardcoded: with the default (no `keep=`), the axes retained are
exactly those the model's own reliability metadata marks trusted, so a
production re-run that promotes gPC2/gPC5 makes the robust score include them
automatically.

### Guarding against extrapolation

A random forest will return a confident-looking number for a site whose climate
lies outside anything in the training data — precisely where it should not be
trusted. `checkExtrapolationRisk()` flags such rows two ways: a per-variable
envelope check (is any predictor outside the training range?) and a Mahalanobis
distance against the training predictor cloud.

```{r deploy-extrap}
er <- checkExtrapolationRisk(mod, newx)
er[, c("mahalanobis", "md_flag", "n_out_of_range", "extrapolation")]
```

On in-domain rows the flag rate is low (about 2–3% at the default 0.975
Mahalanobis quantile); genuinely out-of-domain sites are caught reliably. Treat
a flagged row's prediction as an extrapolation, not an interpolation.

### Chaining to cluster assignment

`scoreThenCluster(..., fit_final = TRUE)` additionally fits and exposes the
final QDA classifier (`final_classifier`, class `geaQDA`). Chaining it after the
score model gives an end-to-end climate → genetic-cluster prediction with a
posterior probability and per-row entropy. The realistic accuracy of this path
is the leakage-free `accuracy` that `scoreThenCluster()` reported — well below
its `ceiling`, because Stage-1 scores are predicted from climate with limited
transferable skill — so the posterior matters as much as the label.

```{r deploy-cluster, eval = FALSE}
s2 <- scoreThenCluster(durumUnits, per_tree = FALSE, fit_final = TRUE,
                       num.trees = 300, min.node.size = 3)
pc <- predictCluster(mod, newx, s2$final_classifier)
pc                                    # predicted cluster, posterior, entropy
```

(This last chunk is shown but not evaluated here purely for vignette build
time; `scoreThenCluster()` with `fit_final = TRUE` re-fits the full Stage-2
model. See `?predictCluster`.)

## Limitations

This workflow, and every number in this vignette, comes from **one** durum
wheat genebank collection (1,060 confident effective units, 43 countries).
The size of the interpolation/extrapolation gap, which targets carry real
within-country signal, and the Stage-2 clustering accuracy and ceiling are
all properties of this dataset. They should not be assumed to generalize to
other crops, other marker panels, or other genebank collections without
re-running this diagnostic pipeline on the new data; only the
*methodology* (pseudoreplication collapsing, spatial-block and
leave-one-group-out CV, fold-safe residualization, permutation significance
testing) is intended to be general-purpose.

Verdicts from `residualize()` and `confoundGapTest()` are based on permutation
p-values; always check the sign and magnitude of the reported R²/gap before
treating a target as carrying real signal, and treat any p-value at or near the
theoretical floor (1/(n_perm+1)) as low-resolution rather than strongly
significant — rerun with a higher `n_perm` to confirm.

## References

Hurlbert, S.H. (1984). Pseudoreplication and the design of ecological field
experiments. *Ecological Monographs*, 54(2), 187-211.

Roberts, D.R. et al. (2017). Cross-validation strategies for data with
temporal, spatial, hierarchical, or phylogenetic structure. *Ecography*,
40(8), 913-929.
